Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-18 icfp.tex
RM edits related & future work

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
We propose a naive $O(n^2)$ algorithm for inferring usage environments of recursive
bindings in Section~\ref{sec:discuss:implementation}, orthogonally to the theory
developed in this section.
%
% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
% environments} used to type each binding.
%
However, we note that inference of usage environments for recursive binding
groups bears some resemblance to the inference of principle types for recursive
bindings traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
% develop a better algorithm leveraging existing inference techniques.
% %
% Despite being a seemingly useful observation, 
We leave the exploration of this connection
as future work.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments Haskell with 
linear types. However, the development is
not concerned with extending GHC's intermediate language(s),
which presents its own challenges. 
%
Nonetheless, while Linear Haskell keeps GHC Core unchanged, its
implementation in GHC does modify and extend Core with linearity/multiplicity
annotations. Core's type system is unable to type \emph{semantic} linearity of
programs in the sense elaborated in our work, which results in Core
rejecting linear programs resulting from optimising transformations that
leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically, in the presence of laziness,
and accepting multiple Core-to-Core passes.
Linear Core can
also be seen as a system that validates the programs written in Linear Haskell
and are compiled by GHC, by guaranteeing (through typing) that linear resources
are still used exactly once throughout the optimising transformations.

@@

Core-to-Core transformations appear in many works across the research
literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
certain optimisations (in particular, let-floating and inlining) greatly
benefit from linearity analysis and, in order to improve those transformation,
special purpose linear-type-inspired systems were created and implemented.
%
By fully supporting linear types in Core, these optimising transformations
could be informed by the language's intrinsic linearity, and, consequently, avoid
an ad-hoc or incomplete linear type inference pass that is custom-built for
optimisations. Additionally, the linearity information may potentially be used
to the benefit of optimising transformations that currently do not take 
linearity into account.

@@

\paragraph{Multiplicity Coercions.}
Linear Core does not have type equality coercions, a flagship feature of GHC
Core's type system.
%
Coercions %briefly explained in Section~\ref{sec:core},
allow the Core intermediate language to encode a panoply of Haskell source
type-level features such as GADTs, type families or newtypes.
%
In Linear Haskell, multiplicities are introduced as annotations to function
arrows which specify the linearity of the function. In practice,
multiplicities are simply types of kind \incode{Multiplicity}, where \incode{One} and \incode{Many}
are the type constructors of the kind \incode{Multiplicity}; multiplicity polymorphism
follows from type polymorphism, where multiplicity variables are
just type variables. Encoding multiplicities as types allows Haskell programs
to leverage features available for types to naturally extend to multiplicities
as well.
For instance, it is possible via the use of GADTs to define a
function whose linearity of its second argument depends on the value
of its first argument, internally realized through so-called
\emph{multiplicity coercions}. Currently, Core cannot make use of such
coercions to determine whether the usages of linear resources match
their intended multiplicity. Studying the interaction between coercions and multiplicities is a main avenue
of future work for Linear Core.
%
% Consequently, we might define, e.g., using a GADT \incode{SBool} and a type family
% |If|, the function \incode{dep} which is linear in the second argument if the first
% argument is \incode{STrue} and unrestricted otherwise:
% %
% \begin{limitation}
% \begin{lstlisting}
% data SBool :: Bool -> Type where
%   STrue :: SBool True
%   SFalse :: SBool False

@@

\paragraph{Optimisations leveraging linearity.}
We only briefly mentioned how linearity can inform optimisations to produce
more performant programs. We leave exploring optimisations unblocked by
preserving linearity in the intermediate language with Linear Core as future
work. Linearity influenced optimising transformations have been also discussed
by Linear Haskell~\cite{cite:linearhaskell} and
in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{cite:linearhaskell}. Additionally, we argue that in Linear
Core accepting more programs as linear there are more chances to use linearity,
in contrast to a linear type system which does not account for lazy evaluation
and thus rejects more programs.

\paragraph{Usage environment resource inference.}
In Section~\ref{sec:linearity-semantically}, we explained that the linear
resources used by a group of recursive bindings aren't obvious and must be
consistent with each other (i.e. considering the mutually-recursive calls) as
though the resources used by each binder are the solution to a set determined
by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
further likened the challenge of determining usage environments for a recursive
group of bindings to a unification problem as that solved by the Hindley-Milner
type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
constraints. Even though these are useful observations, our implementation of
Linear Core uses a naive algorithm to determine the usage environments, thereby
leaving as future work the design of a principled algorithm to determine the
usage environments of recursive group of bindings.

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must account
for all other Core features (e.g. strict constructor fields) and more
optimisations. Despite our initiative in this
direction\footnote{[Link removed due to anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218}
, we leave
this as future work.

>>>>

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
Our implementation uses a naive $O(n^2)$ algorithm for inferring usage
environments of recursive bindings.
% , which is not relevant to the theory developed in this section.
%
% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
% environments} used to type each binding.
%
Moreover, we note that inference of usage environments for recursive binding
groups bears some resemblance to the inference of principle types for recursive
bindings traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
% develop a better algorithm leveraging existing inference techniques.
% %
% Despite being a seemingly useful observation, 
We leave the exploration of this connection
as future work.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
\emph{semantic} linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by proving linearity-preserving multiple Core-to-Core
optimisations GHC.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.

@@

Core-to-Core transformations appear in many works across the research
literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
certain optimisations (in particular, let-floating and inlining) greatly
benefit from linearity analysis and, in order to improve those transformation,
special purpose linear-type-inspired systems were created and implemented.
%
Preserving linear types in Core throughout the compilation pipeline allows the
optimiser to leverage non-heuristic linearity information where, previously, it
would resort solely to ad-hoc or incomplete custom-built linearity inference
passes (naturally, these passes are still necessary to optimise programs not
using explicit linearity or multiplicity annotations). Linearity may
potentially be used to the benefit of other optimising transformations that
currently do not take it into account.

@@

\paragraph{Multiplicity Coercions.}
Linear Core does not have type equality coercions, a flagship feature of GHC
Core's type system.
%
Coercions %briefly explained in Section~\ref{sec:core},
allow the Core intermediate language to encode a panoply of Haskell source
type-level features such as GADTs, type families, or newtypes.
%
In Linear Haskell, multiplicities are introduced as annotations to function
arrows which specify the linearity of the function. In practice,
multiplicities are simply types of kind \incode{Multiplicity}, where \incode{One} and \incode{Many}
are the type constructors of the kind \incode{Multiplicity}; multiplicity polymorphism
follows from type polymorphism, where multiplicity variables are
just type variables. Encoding multiplicities as types allows Haskell programs
to leverage features available for types to naturally extend to multiplicities
as well.
For instance, it is possible via the use of GADTs to define a
function whose linearity of its second argument depends on the value
of its first argument, internally realized through so-called
\emph{multiplicity coercions}. Currently, Core cannot make use of such
coercions to determine whether the usages of linear resources match
their intended multiplicity. Studying the interaction between coercions and multiplicities is a main avenue
of future work for Linear Core.
%
% Consequently, we might define, e.g., using a GADT \incode{SBool} and a type family
% |If|, the function \incode{dep} which is linear in the second argument if the first
% argument is \incode{STrue} and unrestricted otherwise:
% %
% \begin{limitation}
% \begin{lstlisting}
% data SBool :: Bool -> Type where
%   STrue :: SBool True
%   SFalse :: SBool False

@@

\paragraph{Optimisations leveraging linearity.}
We only briefly mentioned how linearity can inform optimisations to produce
more performant programs. We leave exploring optimisations unblocked by
preserving linearity in the intermediate language with Linear Core as future
work. Linearity influenced optimising transformations have been also discussed
by Linear Haskell~\cite{cite:linearhaskell} and
in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{cite:linearhaskell}.
% Additionally, we argue that in Linear Core accepting more programs as linear
% there are more chances to use linearity, in contrast to a linear type system
% which does not account for lazy evaluation and thus rejects more programs.

% \paragraph{Usage environment resource inference.}
% In Section~\ref{sec:linearity-semantically}, we discussed how there is not an obvious way to determine the set of linear
% resources used by a group of (mutually) recursive bindings, .
% though the resources used by each binder are the solution to a set determined
% by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
% further likened the challenge of determining usage environments for a recursive
% group of bindings to a unification problem as that solved by the Hindley-Milner
% type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
% constraints. Even though these are useful observations, our implementation of
% Linear Core uses a naive algorithm to determine the usage environments, thereby
% leaving as future work the design of a principled algorithm to determine the
% usage environments of recursive group of bindings.

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.
